---
title: "Homework 3"
output: pdf_document
---

1. 
A. The posterior is given by $\frac{P(y|\theta)P(\theta)}{\int_{\theta}P(y|\theta)P(\theta)} = \frac{\prod_y (1 + (y - \theta) ^ 2) ^ {-1}}{\int_0^1 \prod_y (1 + (y - \theta) ^ 2) ^ {-1}}$. I computed the integral numerically to get 0.003391051 so we get the following curve 

```{r}
post <- function (th){
     y <- c(-2, -1, 0, 1.5, 2.5)
     p <- prod (cauchy_lik (y, th))/0.003391051
     p}
cauchy_lik <- function (y, th){
  cl <- (1 + (y - th)^2)^(-1)
  cl}
post <- Vectorize(post)
curve(post, 0, 1, xname = "theta")
```

B. $\log P(\theta | y) =$constant $- \sum_y \log(1 + (y - \theta) ^ 2)$

So the first derivative is $\sum_y \frac{2(y - \theta)}{1 + (y - \theta) ^ 2}$

And the second derivative is (using the product rule) $\sum_y \frac{2(y - \theta)^2 - 2}{(1 + (y - \theta)^2)^2)}$

C. We use Newton's method and start with theta = 0. First derivative at 0 is -.18 and the second is -1.32 giving us a posterior mode of -.136

D. The second derivative at the mode is -1.37 so the observed information is 1.37. Therefore the Normal approximation is (-.18, 0.73)
```{r, warning=FALSE}
curve(dnorm, mean=0.18, sd=0.73, 0, 1, xname = "theta")
```
This seems slightly less heavy-tailed than the exact posterior

2.
A. We fit several models in stan. We assumed that the mean of each difference $\text{mean}_j \sim \mathcal{N}(\text{treat}_j, \sigma_j$ where $\sigma_j = (N_j * \text{SE}_j)^2$ We first fit a model where there was a noninformative prior over each $\text{treat}_j$
```{r engine='cat',engine.opts=list(file = "chickens1.stan", lang = "stan")}
data {
  int<lower = 0> J;
  int hz[J];
  int S_N[J];
  int E_N[J];
  vector[J] S_mean;
  vector[J] E_mean;
  vector[J] S_se;
  vector[J] E_se;
}
parameters {
  real treat_effect[J];
  real a;
}
transformed parameters {
  vector[J] sigma;
  vector[J] std;

  for (j in 1:J) {
    std[j] <- (sqrt(E_N[j]) * E_se[j]);
    sigma[j] <- pow(std[j], 2);
  }
}
model {
  for (j in 1:J) {
    E_mean[j] ~ normal(treat_effect[j], sigma[j]);
  }
}
```
```{r warning=FALSE, message=FALSE, results='hide'}
library ("rstan")
setwd("~/Documents/BDA/Homework 3")
chickens <- read.table("chickens_data.txt", header=TRUE)
hz <- chickens$Hz
S_N <- chickens$S_N
E_N <- chickens$E_N
S_mean <- chickens$S_Mean
E_mean <- chickens$E_Mean
S_se <- chickens$S_SE
E_se <- chickens$E_SE
J <- length(hz)

fit1 <- stan("chickens1.stan")
```
From this we get that the treatment effect at each level of electromagnetic activity was fairly similar.
```{r}
plot(fit1, pars = "treat_effect")
```
B. So we then tested a model that put a prior on the treat effects to see if they were generated from the same distribution. We used $\text{treat}_j \sim \mathcal{N} (1, 1)$
```{r engine='cat',engine.opts=list(file = "chickens2.stan", lang = "stan")}
data {
  int<lower = 0> J;
  int hz[J];
  int S_N[J];
  int E_N[J];
  vector[J] S_mean;
  vector[J] E_mean;
  vector[J] S_se;
  vector[J] E_se;
}
parameters {
  real treat_effect[J];
}
transformed parameters {
  vector[J] sigma;
  vector[J] std;

  for (j in 1:J) {
    std[j] <- (sqrt(E_N[j]) * E_se[j]);
    sigma[j] <- pow(std[j], 2);
  }
}
model {
  for (j in 1:J) {
    treat_effect[j] ~ normal(1, 1);
    E_mean[j] ~ normal(treat_effect[j], sigma[j]);
  }
}
```
```{r warning=FALSE, message=FALSE, results='hide'}
fit2 <- stan("chickens2.stan")
```
This didn't seem to affect the fit much at all suggesting that the treatment effects were all drawn from a similar distribution
```{r}
plot(fit2, pars = "treat_effect")
```
C. The sham treatment ensures that the effect is not simply due to being in air rather than water.
D. We fit a third model where we modeled the treatment effects for both the sham and the shocked chicken brains (using the same model). This actually improved the fit, going from a log posterior of 90 to 186.
```{r engine='cat',engine.opts=list(file = "chickens3.stan", lang = "stan")}
data{
  int<lower = 0> J;
  int hz[J];
  int S_N[J];
  int E_N[J];
  vector[J] S_mean;
  vector[J] E_mean;
  vector[J] S_se;
  vector[J] E_se;
}
parameters {
  real treat_effect_E[J];
  real treat_effect_S[J];
}
transformed parameters {
  vector[J] sigma_E;
  vector[J] sigma_S;
  vector[J] std_E;
  vector[J] std_S;

  for (j in 1:J) {
    std_E[j] <- (sqrt(E_N[j]) * E_se[j]);
    sigma_E[j] <- pow(std_E[j], 2);
    std_S[j] <- (sqrt(S_N[j]) * S_se[j]);
    sigma_S[j] <- pow(std_S[j], 2);
  }
}
model {
  for (j in 1:J) {
    treat_effect_E[j] ~ normal(1, 1);
    treat_effect_S[j] ~ normal(1, 1);
    E_mean[j] ~ normal(treat_effect_E[j], sigma_E[j]);
    S_mean[j] ~ normal(treat_effect_S[j], sigma_S[j]);
  }
}
generated quantities {
  real treat_effect[J];
  for (j in 1:J) 
    treat_effect[j] <- treat_effect_E[j]/treat_effect_S[j];
}
```
```{r warning=FALSE, message=FALSE, results='hide'}
fit3 <- stan("chickens3.stan")
```
```{r}
plot(fit3, pars = "treat_effect")
```